Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
After reading through everything

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{zhu15, buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, \textbf{methods} may only identify signs of \emph{non}-convergence \citep{hoff09}. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when approximate convergence, necessary for valid inferences, may be safely concluded. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}}

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\) in dataset \(y\). The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\). For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the \textbf{imputed values}). The collection of \(\theta\)-values between \(t=1\) (summarizing the state space of the algorithm at initialization) and \(t=T\) (summarizing the state space for the imputed values) will be referred to as an `imputation chain'.

@@

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain. \textbf{Note that \(B\) refers to the variance in \(\theta\), instead of the typical notation in MI where \(B\) denotes the variance between estimated quantities of scientific interest \(\hat{Q}_{\ell}\).}

@@

When we look at panel B, we conclude something weird. The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter. Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on \texttt{stats::acf()}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms. The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower. Therefore, we compute \(AC\) manually, see the solid line. These \(AC\) values indicate non-stationarity as expected. We therefore calculate \(AC\) manually.

\textbf{Describe panel C here and why it is not optimal: NB. In panel C: De grotere variantie overall zorgt er kennelijk voor dat \(\widehat{R}\) groeit. De \(\widehat{R}\) geeft de foutieve boodschap dat convergentie steeds slechter gaat. Verklaring: de overdispersie assumptie klopt niet als er trend is.{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above. \(\widehat{R}\) as computed conform \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up. \textbf{\citep[explain the dip in Rhat values at t=2. Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4). That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}}

\textbf{In conclusion, we will use \(\widehat{R}\) conform \citet{veht19} and \(AC\) computed manually as non-convergence diagnostics in the simulation study. The diagnostics will be applied to four \(\theta\)s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix. Results in eight methods to identify non-convergence}

@@

The aim of the simulation study is to determine the \textbf{impact of non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in iterative imputation algorithms. Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm. \textbf{{[}REMOVE?{]} Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence }{[}and performance of different thetas{]}\textbf{. And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice. That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\).}

To induce non-convergence, we consider two sets of simulation conditions: we vary the missingness proportion \(p_{\rm mis}\) in the incomplete data, and we vary the number of iterations \(T\) in the imputation algorithm. The missingness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions should yield slower convergence. Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data. Then, convergence to a stable but highly variable state may be reached instantaneously. We expect that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.

The assumption inherent to the second set of simulation conditions---the number of iterations---is that terminating the imputation algorithm too early causes non-convergence. Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm (which are sampled randomly from the set of observed datapoints). As the number of iterations increases, the imputation chains should become independent of the initial values, until the point at which adding an extra iteration does not lead to a more converged state. We assume that we can induce non-convergence at least in conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness proportion \(p_{\rm mis}\) and a low number of iterations \(T\) will more often result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence-valid estimates of the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}. We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\). \textbf{{[}ADD limitation that other MI packages may have different way to obtain initial values!{]}}
\item
  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  We further hypothesize that multivariate \(\theta\)s are better at detecting non-convergence than univariate \(\theta\)s.
\end{enumerate}

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions \citep{vink14}. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem. Let the predictor space be defined as
\[
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 \\
4 & 16 & 4.8\\
1.8 & 4.8 & 9
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
\]
A finite population of \(N=1000\) is simulated using the \texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth variable is constructed as outcome variable \(Y\). For each unit \(i = 1, 2,..., N\), let
\[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\]
where \(\epsilon \sim \mathcal{N}(0, 100)\). This results in a complete set \(y\), with observed values for all units \(i\) on all variables \(j\) (where \(j = Y, X_1, X_2, X_3\)). From the complete set we obtain the true values of the scientific estimands \(Q\).

@@

We consider four types of \(Q\)s that are often of interest in empirical research. Namely, two descriptive statistics: the mean \(\mu_j\) and standard deviation \(\sigma_j\) of each variable \(j\). We also consider the regression coefficients of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in regression equation
\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\]
where \(Y'\) is the expected value of the outcome. And finally, the proportion of variance explained by the set of predictors: coefficient of determination \(r^2\) \textbf{(note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\))}. These \(Q\)s are estimated by \(\bar{Q}\) in each simulation condition.

@@

The complete dataset is \emph{amputed} in each simulation repetition to create five incomplete datasets \(y\) with different missingness proportions \(p_{\rm mis}\). The \(p_{\rm mis}\) varies between 5 and 95\%, in incremental steps (\(p_{\rm mis}\) = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep{rubin87}. I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{\rm mis}\). This is performed with the `mice' package \citep[function \texttt{mice::ampute()};][]{mice}.

The second set of simulation conditions is the number of iterations in the imputation algorithm. We vary the number of iterations between one and one hundred (\(T = 1,2,.,100\)). All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(m=5\)). We perform this with the \texttt{mice()} function \citep{mice}. The result is five sets of imputations per simulation condition, \(y_{\rm imp, \ell}\).

Subsequently, we obtain the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\} by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}. Univariate quantities of scientific interest \(Q\) are estimated directly from the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. Estimates of multivariate \(Q\)s are obtained after employing multiple linear regression \citep[function \texttt{stats::lm()},][]{R}. From the estimates in each imputation, \(\hat{Q}_\ell\), we obtain the pooled estimate \(\bar{Q}\). For univariate estimates we pool \(\hat{Q}_\ell\) as average. Multivariate \(\hat{Q}_\ell\) are pooled conform \citet{vink14}.

@@

In total, we consider eight methods to identify non-convergence. Namely, we apply the two non-convergence diagnostics \(\widehat{R}\) and \(AC\) on four different \(\theta\)s. Univariate \(\theta\)s are obtained from \(y_{\rm imp, \ell}\). The multivariate \(\theta\) \(\lambda_{1,\ell}\) is obtained from \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. The other multivariate \(\theta\) is equal to the estimated regression coefficient \(\hat{Q}_\ell\) in \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. We apply the two non-convergence diagnostics to each \(\theta\) by implementing \citet{veht19} 's recommended \(\widehat{R}\) and we compute \(AC\) manually.

@@

As recommended by \citet{buur18} we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW). Bias is calculated for each \(Q\). CR and CIW only for \(Q = \beta\).
We calculate bias as \(\bar{Q} - Q\). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\). Let
\[\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},\]
where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(m-1\) degrees of freedom, and \(\rm SE_{\bar{Q}}\) is the square root of the pooled variance estimate. If we obtain nominal coverage (CR = 95\%), we can conclude that \(\bar{Q}\) is a confidence-valid estimate of \(Q\). Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\). CIW is of interest because it is a measure of efficiency. Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.

We evaluate the methods to identify non-convergence against these performance measures. \(\widehat{R}\) and \(AC\) should identify non-convergence in simulation conditions, where \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).

@@

\textbf{{[}Add more info about figure legends and axes{]}}

Our results show the performance of the MICE algorithm under different conditions of missingness and iteration length. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands (\(Q=\mu\) and \(Q=\sigma\)) we observe the largest bias in the outcome variable \(Y\). And for the regression coefficients (\(Q=\beta\)) the bias is most pronounced in \(\beta_1\) (the effect of \(X_1\) on \(Y\)). Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\). In the figures, we present all missingness conditions, but only \textbf{iteration length conditions} \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \geq 30\). Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

Figure \ref{fig:Qs} displays the influence of \textbf{inducing non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} on the estimated quantities of scientific interest, \(Q\)s. \textbf{{[}Add panel labels, describe panels{]}} Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.

@@

The estimated univariate means seem unaffected by the number of iterations in the algorithm. This implies that approximately unbiased estimates may be obtained with as little as one iteration (\(T \geq 1\)). The magnitude of the bias in \(\bar{Q}\) depends solely on the missingness proportion \(p_{\rm mis}\), with missingness proportions of 75\% and 95\% leading to more extreme biases than other missingness proportions. Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}\leq.50\). \textbf{This is curious because the missingness mechanism MCAR should yield unbiased univariate estimates without employing multiple imputation.} Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)). We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).

@@

As Figure \ref{fig:rhats-acs} shows, the results of the two diagnostics applied to the chain variances are highly similar to those on the chain means. This holds for both \(\widehat{R}\) and \(AC\). The only apparent difference is in the \(\widehat{R}\) threshold 1.1. Instead of diagnosing non-convergence in conditions where \(T\leq13\), we now diagnose this when \(T\leq11\). All other observations are equivalent.

>>>>

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{zhu15, buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of \emph{non}-convergence \citep{hoff09}. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged \emph{enough} for valid inferences. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}}

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the proportion of incomplete cases \(p_{\rm inc}\) in dataset \(y\). The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\). For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the \textbf{imputed values}). The collection of \(\theta\)-values between \(t=1\) (summarizing the state space of the algorithm at initialization) and \(t=T\) (summarizing the state space for the imputed values) will be referred to as an `imputation chain'.

@@

\noindent We then estimate between-chain variance \(B\) (note that we diverge from the typical notation in MI, where \(B\) denotes the variance between the estimated quantities of scientific interest \(\hat{Q}_{\ell}\)). \(B\) is defined as the variance of the collection of average \(\theta\)s per chain:

@@

When we look at panel B, we conclude something weird. The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter. Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\) for both scenarios. However, if we look at the chain means of the non-convergence scenario, there should be signs of trending up-to iteration number seven. After consulting the documentation on \texttt{stats::acf()}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms. The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower. Therefore, we compute \(AC\) manually, see the solid line. These \(AC\) values behave as expected from visual inspection. We will, therefore, only consider manually calculated \(AC\)-values as non-convergence diagnostic.

\textbf{Describe panel C here and why it is not optimal: De grotere variantie overall zorgt er kennelijk voor dat \(\widehat{R}\) groeit. De \(\widehat{R}\) geeft de foutieve boodschap dat convergentie steeds slechter gaat. Verklaring: de overdispersie assumptie klopt niet als er trend is.{]}} The \(\widehat{R}\)-values in panel D are conform our expectations. The adapted version of \(\widehat{R}\) proposed by \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up. \textbf{\citep[explain the dip in Rhat values at t=2. Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4). That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}} \newline

\noindent In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms. We will compute \(\widehat{R}\) conform \citet{veht19}, and calculate \(AC\) manually. The diagnostics will be applied to the four types of scalar summaries \(\theta\) described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix. We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four \(\theta\)s) through simulation.

@@

The aim of the simulation study is to determine to what extend non-convergence affects iterative imputation algorithms. Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm. And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods. We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest \(\bar{Q}\) are unbiased and confidence-valid estimates of \({Q}\). In contrast to this, we defined \emph{approximate} convergence as the most converged state that the algorithm can obtain. Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving \(\widehat{R}\)- and \(AC\)-values). Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.

To induce non-convergence, we consider two sets of simulation conditions: `incompleteness' and `early stopping'. Incompleteness refers to the proportion of incomplete cases in dataset \(y\), \(p_{\rm inc}\). Early stopping is defined by the number of iterations in the imputation algorithm, \(T\).

The incompleteness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases \(p_{\rm inc}\) leads to quick algorithmic convergence, since there is a lot of information in the observed data. Higher \(p_{\rm inc}\) yields slower convergence---unless there is so little information in \(y_{\rm obs}\) that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. \textbf{Still, we expect that the incompleteness conditions will result in more signs of non-convergence.}

The assumption inherent to the second set of simulation conditions, early stopping, is that terminating the imputation algorithm too early causes non-convergence. Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm \citep[which are sampled randomly from the set of observed datapoints;][]{buur18}. As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state). We expect that we can induce non-convergence by early stopping in at least those conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\) \citep{mice}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\) will more often result in biased, invalid estimates of the quantities of scientific interest \(Q\).
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where any \(\bar{Q}\) is \emph{not} an unbiased and confidence-valid estimate of \(Q\).
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}. We, therefore, suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  We further hypothesize that multivariate \(\theta\)s are better at detecting non-convergence than univariate \(\theta\)s.
\end{enumerate}

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions \citep{vink14}. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem. Let the predictor space be defined as
\[
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 \\
4 & 16 & 4.8\\
1.8 & 4.8 & 9
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
\]
A finite population of \(N=1000\) is simulated using the \texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth variable is constructed as outcome variable \(Y\). For each unit \(i = 1, 2,..., N\), let
\[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\]
where \(\epsilon \sim \mathcal{N}(0, 100)\). This results in a dataset \(y\), where \(y_{\rm mis} = \emptyset\). In other words, \(y\) contains the complete set of observations for all units \(i\) on all variables \(j\) (where \(j = Y, X_1, X_2, X_3\)). From the complete dataset we obtain the true values of the scientific estimands \(Q\).

@@

We consider four types of \(Q\)s that are often of interest in empirical research. Namely, two descriptive statistics: the mean \(\mu_j\) and standard deviation \(\sigma_j\) of each variable \(j\). We also consider the regression coefficients of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in regression equation
\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\]
where \(Y'\) is the expected value of the outcome. The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination \(r^2\) (note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\)). The scientific estimands are estimated by \(\bar{Q}\) in each simulation condition.

@@

In each simulation repetition, we \emph{ampute} the complete dataset conform the `incompleteness' conditions. The proportion of incomplete cases \(p_{\rm inc}\) varies between 5\% and 95\%, in incremental steps (\(p_{\rm inc}\) = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep[MCAR;][]{rubin87}. I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on \(p_{\rm inc}\). We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five `incompleteness' conditions.

We impute the missing values in \{\(y_{\rm obs}, y_{\rm mis}\)\} according to the `early stopping' conditions. That is, we vary the number of iterations in the imputation algorithm between one and one hundred (\(T = 1,2,.,100\)). All imputation procedures are performed using \texttt{mice} \citep[function \texttt{mice()};][]{mice}, with Bayesian linear regression imputation, and five imputation chains (\(m=5\)). This results in five sets of imputations for each of the one hundred `early stopping' conditions, \(y_{\rm imp, \ell}\).

Subsequently, we obtain the completed data per imputation, \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}, by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}. We estimate univariate \(Q\)s directly from the completed data, by averaging across imputations to obtain \(\bar{Q}\). Estimates of multivariate \(Q\)s are obtained after performing multiple linear regression \citep[function \texttt{stats::lm()},][]{R}. We pool the resulting estimates, \(\hat{Q}_\ell\), into a single \(\bar{Q}\) conform \citet{vink14}.

@@

We consider eight sets of diagnostic methods to identify non-convergence, by applying \(\widehat{R}\) and \(AC\) on four different \(\theta\)s. Univariate \(\theta\)s are obtained from \(y_{\rm imp, \ell}\). The model-independent multivariate \(\theta\), \(\lambda_{1}\), is obtained from \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. The model-dependent multivariate \(\theta\) is one of the quantities of scientific interest: the estimated regression coefficients in \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. We apply the two non-convergence diagnostics on the \(\theta\)s by implementing the adapted version of \(\widehat{R}\) and by manually programming the \(AC\) function.

@@

As recommended by \citet{buur18}, we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW). Bias is calculated for each \(Q\), whereas CR and CIW are only obtained for \(Q = \beta\). We calculate bias as \(\bar{Q} - Q\). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\). Let
\[\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},\]
where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(m-1\) degrees of freedom, and \(\rm SE_{\bar{Q}}\) is the square root of the pooled variance estimate. If we obtain nominal coverage (CR = 95\%), we can conclude that \(\bar{Q}\) is a confidence-valid estimate of \(Q\). Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\). CIW is of interest because it is a measure of efficiency. Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.

We evaluate the diagnostic methods to identify non-convergence against these performance measures. \(\widehat{R}\) and \(AC\) should identify non-convergence in simulation conditions, where \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).

@@

Our results show the performance of the MICE algorithm under different conditions of `incompleteness' and `early stopping'. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands (\(Q=\mu\) and \(Q=\sigma\)) we observe the largest bias in the outcome variable \(Y\). And for the regression coefficients (\(Q=\beta\)) the bias is most pronounced in \(\beta_1\) (the effect of \(X_1\) on \(Y\)). Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\). In the figures, we present all `incompleteness' conditions, but only `early stopping' conditions where \(1 \leq T\leq50\). That is, the results are more or less stable for conditions where \(T \geq 30\). Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.

@@

Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, \(Q\)s. \textbf{{[}Add panel labels, describe panels{]}} Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.

@@

The estimated univariate means seem unaffected by early stopping. This implies that approximately unbiased estimates may be obtained with as little as one iteration (\(T \geq 1\)). The bias in \(\bar{Q}\) depends solely on the proportion of incomplete cases, with \(p_{\rm inc}\geq.75\) leading to more extreme biases than other conditions. Completely unbiased estimates are only obtained in conditions where \(p_{\rm inc}\leq.50\). \textbf{This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.} Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)). We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).

@@

As panel C and D in Figure \ref{fig:rhats-acs} show, the results of this set of methods are highly similar to those with chain means as \(\theta\). This holds for both \(\widehat{R}\) and \(AC\). The only apparent difference is in the \(\widehat{R}\) threshold 1.1. Instead of diagnosing non-convergence in conditions where \(T\leq13\), we now diagnose this when \(T\leq11\). All other observations are equivalent.

\hypertarget{thetahatq.}{%
\subsubsection{\texorpdfstring{\(\theta=\hat{Q}\).}{\textbackslash theta=\textbackslash hat\{Q\}.}}\label{thetahatq.}}
